---
title: "Bootstrapping"
author: "Caelan Bryan, Jenna Dufresne, Jamielee Jimenez Perez"
date: "Fall 2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
course: STA 6257 - Advance Statistical Modeling
bibliography: references.bib # file contains bibtex for references
always_allow_html: true # this allows to get PDF with HTML features
nocite: | 
  @efron1
  @efron2
  @introduction
  @assumptions
  @withr
  @chavez_2022
  @worldbank
---

## Literature Review

Bootstrap Methods: Another Look at the Jackknife [@efron1] is credited as being 
the creation of bootstrapping by author B. Efron. It starts by explaining that
bootstrapping is a step above the jackknife method, and that the jackknife is
a linear expansion for estimating the bootstrap. Starting in the introduction,
it is stated that the bootstrap is shown to estimate the variance of the sample
median, which is an area that the jackknife fails at. It is also mentioned that
the bootstrap does well at estimating the error rates in certain problems, which
outperforms other non-parametric estimation methods. The problem attempting to
be solved is estimating the sampling distribution based on the observed data.
The idea behind the bootstrap method is listed in three parts. First, construct
the sample probability distribution. Second, draw a random sample of size n. 
Lastly, approximate the sampling distribution by the bootstrap distribution. It
is mentioned that that the difficult part of bootstrapping is calculating the 
bootstrap distribution and three methods are given to accomplish this; direct
Monte Carlo approximations, and Taylor series expansion methods. A few
applications are listed. These include estimating the median, error rate 
estimation in discrimination analysis, relationship with the jackknife,
Wilcoxon's statistic, and regression models. Finally, a list of some remarks
regarding the bootstrap method are listed. Some important ideas listed are that
the calculation of the bootstrap distribution using the Monte Carl method is
easy to implement on the computer and that the bootstrap and jackknife provide
approximate frequency statements and not approximate likelihood statements.

Nonparametric Estimates of Standard Error: The Jackknife, the Bootstrap, and
Other Methods [@efron2] starts by giving some background on what the purpose of 
the paper is. In this case, we want to estimate the standard error based on the 
data. This is normally done through parametric modeling methods, but in the 
paper it is being done through nonparametric methods. All methods are being 
tested using the same scenario which is bivariate normal distribution. This 
paper has four points. They are to describe the various methods, show how the 
methods derive from the same idea, relate the methods to each other, and finally 
show how the methods perform differently even though they are similar. The Monte 
Carlo experiment that the data was obtained from is then described. The 
bootstrap method is the first that is looked at. The steps are described to 
obtain the bootstrap samples. Random samples are created from the original 
sample and the bootstrap estimate is obtained. This process is than repeated 
multiple time to achieve enough samples to accurately estimate the standard 
error. 128 and 512 total samples were used separately, and while the 512 total 
sample provided a slightly more accurate result, the increase was minimal. 
on this result, the size of N samples is not overly important past 50-100. A 
smoothed bootstrap method is than looked at. This produced by compromising the 
normal theory maximum, likelihood estimate and the nonparametric maximum 
likelihood estimate. The results of the smoothed bootstrap method were overall 
better than those in the on-smoothed bootstrap method. Next, the jackknife, 
infinitesimal jackknife, half-sampling, and random sub sampling methods are 
discussed. All results are displayed in a table containing th estimated values, 
standard deviation, and confidence interval for each estimate. Overall, the 
bootstrap method produced the results that closest matches the theoretical 
values.

In Bootstrapping - An Introduction And Its Applications In Statistics 
[@introduction] the authors take a high-level approach to introducing the reader 
to bootstrapping along with applications to different parts of statistics. The 
introduction begins to explain what bootstrapping is and how it is accomplished. 
They make use of an example to help explain what bootstrapping is. In this case, 
they look at performing analysis on the population of the United States. It 
would be difficult, costly, and timely to sample the entire population so 
instead you might sample a smaller subset of the population and create new 
bootstrap samples using replacement. In this sense the bootstrap samples might 
contain duplicates or even omit certain responses from the original sample. 
Doing so allows interpretation of the entire population based on a subset of the 
population. One of the first applications looked at is estimation of means and 
confidence intervals for the mean. Another application is constructing 
confidence intervals for regression coefficients. The authors also look at 
regression models, case re-sampling, estimating the distribution of sample mean, 
Bayesian bootstrap, smooth bootstrap, parametric bootstrap, and re-sampling 
residuals as applications of bootstrapping. Lastly, some of the advantages and 
disadvantages are listed.  One advantage is the simplicity of deriving estimates 
of standard errors and confidence intervals where a disadvantage would be that 
the result still depends on the original sample.

In The Importance of Discussing Assumptions when Teaching Bootstrapping
[@assumptions] the authors spend time ensuring that the readers understand when 
it is appropriate to use bootstrapping methods by presenting the math along with 
experimental results. The paper starts with an introduction to what 
bootstrapping is. They also mention that bootstrapping has increased in 
popularity since it’s inception with applications in linear regression and 
neural networks among other fields. The methods focused on during the paper are 
studentized, basic, and percentile bootstrap intervals and their hypothesis 
tests. The next section focuses on why it’s important to teach statistical 
computing and bootstrapping. One feature relevant to bootstrapping is that 
students understandings of confidence intervals and statistical inference relies 
on their understanding of sampling distributions. Next, some of the assumptions 
for bootstrapping are discussed. The assumptions were split based on interval 
estimation and hypothesis testing. For both cases, the largest assumption is 
that the distribution can be made approximately pivotal through shifting or 
studentization. Lastly, simulation-based performance was evaluated. This was 
done by using the metrics coverage proportion, significance level, and power. 
The most important results were that when the assumptions are broken, there can 
be differences in the performance of the different bootstrapping methods. There 
also was not a improvement between bootstrapping and other methods whose 
assumptions were also broken. The smaller the sample size and non-normalcy also 
impacted the performance of the methods. Lastly, an R package was created, which
encompasses the functions used throughout the paper, that can be used to create 
intervals using bootstrapping methods.

Bootstrapping with R to make generalized inference for regression model [@withr] 
looks at a specific application of bootstrapping, validating a generalized 
regression model to make generalization of statistical inference to different 
cases outside of the original sample. In the introduction, the authors explain 
the different types of regression models and explain the different ways to 
complete model validation. Some of those listed include cross-validation, 
Jackknife, and bootstrap methods. The idea of the bootstrap method is then also 
explained. The methodology and design of the experiment is then outlined. In 
this case, the original observations will be resampled leading to a set of 
bootstrap samples. The mean estimate and regression coefficient estimates can 
then be found for each bootstrap sample. Finally, confidence intervals can be 
created for the mean, regression coefficients, and standard errors for both the 
mean and regression coefficients. A table is provided to outline the values 
received from the original sample and bootstrap samples. Finally, the results 
are discussed in the conclusion. Based on the results of this experiment, the 
values found are very similar among the original and bootstrap samples, although 
the confidence intervals for the bootstrap samples are often wider. Some 
advantages to bootstrapping are also listed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(knitr)
```

## Data Description

The dataset we will be using to explore Bootstrapping is the "Causes of Death -
Our World in Data" dataset from kaggle [@chavez_2022]. The causes of death 
dataset was also expanded with population data from the world bank [@worldbank].

The raw data contains a multitude of death statistics broken down by the 
continent, region, country, and territory. The statistics for cause of death
are given in number of deaths and split up by the cause of death. Data is
presented over the span of multiple years. The population dataset from the
world bank is a list of countries, regions, and territories but contains the
population for each year from 1960 to 2021 where available.

To get more meaningful numbers, the population for each year 1990 through 2019
was populated into the causes of death dataset. In order to perform a more
predictable experiment, the causes of death dataset was cleaned by first removing
all non-country entries. The number of executions and terrorism deaths columns
were also removed due to a lack of data for the majority of countries. As a
final note, the Vatican and Liechtenstein are the two countries that did not
have cause of death statistics available in the dataset.

The columns in the cleaned dataset are as follows:

```{r echo=FALSE}
Name <- c('Entity', 
          'Population', 
          'Code', 
          'Year', 
          'Deaths_Meningitis', 
          'Deaths_Neoplasms',
          'Deaths_FireHeatHotSubstances', 
          'Deaths_Malaria', 
          'Deaths_Drowning', 
          'Deaths_InterpersonalViolence', 
          'Deaths_HIVAIDS', 
          'Deaths_DrugUseDisorders',
          'Deaths_Tuberculosis', 
          'Deaths_RoadInjuries', 
          'Deaths_MaternalDisorders', 
          'Deaths_LowerRespiratoryInfections',
          'Deaths_NeonatalDisorders',
          'Deaths_AlcoholUseDisorders',
          'Deaths_ExposureToForcesOfNature',
          'Deaths_DiarrhealDiseases',
          'Deaths_EnvironmentalHeatAndColdExposure',
          'Deaths_NutritionalDeficiencies',
          'Deaths_Selfharm',
          'Deaths_ConflictAndTerrorism',
          'Deaths_DiabetesMellitus',
          'Deaths_Poisonings',
          'Deaths_ProteinEnergyMalnutrition',
          'Deaths_CardiovascularDiseases',
          'Deaths_ChronicKidneyDisease',
          'Deaths_ChronicRespiratoryDiseases',
          'Deaths_CirrhosisOtherChronicLiverDiseases',
          'Deaths_DigestiveDiseases',
          'Deaths_AcuteHepatitis',
          'Deaths_AlzheimersDiseaseOtherDementias',
          'Deaths_ParkinsonsDisease')
Description <- c('Name of Country',                    
                 'Population of Country at Specific Year',                    
                 'Three Letter Country Code',                    
                 'Year for Causes of Deaths',              
                 'Number of Deaths Caused by Meningitis',                
                 'Number of Deaths Caused by Neoplasms',         
                 'Number of Deaths Caused by Fire, Heat, or Hot Substances',                
                 'Number of Deaths Caused by Malaria',      
                 'Number of Deaths Caused by Drowning',                
                 'Number of Deaths Caused by Interpersonal Violence',      
                 'Number of Deaths Caused by HIV/AIDS',                
                 'Number of Deaths Caused by Drug Use Disorders',      
                 'Number of Deaths Caused by Tuberculosis',                
                 'Number of Deaths Caused by Road Injuries',      
                 'Number of Deaths Caused by Maternal Disorders',                
                 'Number of Deaths Caused by Lower Respiratory Infections',      
                 'Number of Deaths Caused by Neonatal Disorders',                
                 'Number of Deaths Caused by Alcohol Use Disorders',      
                 'Number of Deaths Caused by Exposure to Forces of Nature',                
                 'Number of Deaths Caused by Diarrheal Diseases',      
                 'Number of Deaths Caused by Environmental Heat and Cold Exposure',                
                 'Number of Deaths Caused by Nutritional Deficiencies',      
                 'Number of Deaths Caused by Self-Harm',                
                 'Number of Deaths Caused by Conflict and Terrorism',      
                 'Number of Deaths Caused by Diabetes Mellitus',                
                 'Number of Deaths Caused by Poisoning',      
                 'Number of Deaths Caused by Protein Energy Malnutrition',                
                 'Number of Deaths Caused by Cardiovascular Diseases',      
                 'Number of Deaths Caused by Chronic Kidney Disease',                
                 'Number of Deaths Caused by Chronic Respiratory Diseases',      
                 'Number of Deaths Caused by Cirrhosis or Other Chronic Liver Diseases',                
                 'Number of Deaths Caused by Digestive Diseases',      
                 'Number of Deaths Caused by Acute Hepatitis',                
                 'Number of Deaths Caused by Alzheimers Disease or Other Dementias',      
                 'Number of Deaths Caused by Parkinsons Disease')                        
Type <- c('Nominal',  
          'Discrete',   
          'Nominal',
          'Nominal',   
          'Discrete',  
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete',
          'Discrete')
variable_data <- data.frame(Name, Description, Type)
kable(variable_data)
```

## Introduction

## Methods

### Statistical Modeling

### Conlusion

## References